---
title: "Tutorial Part II - R goes spatial"
author: "Stephanie Kramer-Schadt"
date: "`r Sys.setlocale('LC_TIME','C'); paste('Last Update', format(Sys.time(), '%B %e, %Y')) `" 
        #"`r Sys.Date()`" # 
output:
  html_document:
    code_folding: show
    link-citations: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '4'
  slidy_presentation: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = 'C:/Users/kramer/PopDynIZW Dropbox/Steph Kramer/_GitHub')
```

# Introduction

Let me start with a big thanks to Moritz Wenzler, Cedric Scherer, Pierre Gras and Juergen
Niedballa who constantly helped to improve the course! I am deeply indepted to you!

The course is intended to be a quick introduction to using R as a GIS. It builds
on the previous course, Day1_RIntro, and is written in such a way that you can 
understand handling spatial data when knowing the basics of handling data.frames, 
matrices and lists. There is no need to know tidyverse/dplyr coding style.

This course provides the basics for using the most important spatial 
data types - vector (points, lines, polygons, aka ESRI shapefiles) and raster data.
In R, these are *Spatial(Points/Lines/Polygons)DataFrame* or *RasterLayer* Objects. 
Recently, *sf*-objects have been developed to handle vector data more easily. The
latter can be treated like simple data.frames. Also the raster package is
under development - called *terra* package - but the course does not cover this.
The *terra* package is especially useful for remote sensing data.

In addition to different data types, the course provides sections on coordinate 
projections and the most important geospatial operations. Last but not least - 
we plot a lot to ease data visualisation and because it is fun.

The data we use stem from a longstanding project we run in Borneo. Please refer to
our website www.ecological-dynamics-izw.com. The project involves species
conservation and large scale landscape planning. To learn more about the
data used in the course and the project, please refer to the following publications 
(freely available):

* Targeted conservation to safeguard a biodiversity hotspot from climate and land-cover change.
MJ Struebig, et al. 2015. Current Biology 25 (3), 372-378

* Anticipated climate and land‐cover changes reveal refuge areas for Borneo's orang-utans.
MJ Struebig, et al. 2015. Global Change Biology 21 (8), 2891-2904

* The importance of correcting for sampling bias in MaxEnt species distribution models.
S Kramer‐Schadt et al. 2013. Diversity and Distributions 19 (11), 1366-1379.

* The Borneo carnivore database and the application of predictive distribution modelling.
S Kramer-Schadt, et al.2016. Raffles Bulletin of Zoology, Supplement No. 33: 18–41.

 
## Useful (web)sites

1. R news and tutorials
* http://www.r-bloggers.com/

2. Quick introduction to spatial R
* https://rspatial.org/
* http://pakillo.github.io/R-GIS-tutorial/
* http://rstudio-pubs-static.s3.amazonaws.com/7993_6b081819ba184047802a508a7f3187cb.html

3. Spatial visualisation
* check out the packages tmap, ggmap, leaflet and cartography
* https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html
* A whole book: https://clauswilke.com/dataviz/  with spatial data here:
  https://clauswilke.com/dataviz/geospatial-data.html

4. overview over spatial packages
* http://cran.r-project.org/web/views/Spatial.html # http://cran.r-project.org/view=Spatial
* https://www.r-spatial.org/  

5. R-cheatsheets - great for learning 'vocabulary'
* in your folder 
* or here: https://www.rstudio.com/resources/cheatsheets/

6. Infos about simple features in R (sf package) 
*  https://r-spatial.github.io/sf/index.html   
*  and nice intro: https://oliviergimenez.github.io/introspatialR/#1
*  geocomputation: https://geocompr.robinlovelace.net/spatial-operations.html

7. Use EPSG codes as unique and specific identifies of your coordinate reference systsem instead of writing projection details. 
   See http://spatialreference.org/ for EPSG codes or
   https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/OverviewCoordinateReferenceSystems.pdf
   
8. BioMove specials:
*  For analysis of  plot-based biodiversity data: package **vegan**
*  For analysis of telemetry data: packages **adehabitatHR, adehabitatLT, move,recurse, momentuHMM, moveHMM, ctmm**
   
## Further reading

1. Check Roger Bivand's publications:
* http://cran.r-project.org/web/views/Spatial.html
* http://www.csiss.org/learning_resources/index.html
* Roger Bivand et al. 2008, Applied Spatial Data Analysis in R, Springer

2. Check Edzer Pebesma's course materials and publications:
* http://ifgi.uni-muenster.de/~epebe_01/
* https://edzer.github.io/UseR2017/  
* https://www.rstudio.com/resources/videos/tidy-spatial-data-analysis/  

3. Geocomputation with R: a free book of Robin Lovelace, Jakub Nowosad, Jannes Muenchow
* https://geocompr.robinlovelace.net

4. DataVizzArt  Cedric Scherer -> check out his #TidyTuesday or #30DaysMapChallenge contributions
* https://www.cedricscherer.com/top/dataviz/


# Basics

## How to install...

```{r}
pkgs = c("here", "sp", "sf", "dismo", "raster", "GISTools",  "rgdal", 
         "maptools", "rgeos","rgl","rasterVis", "ellipse", "units", "tmap", 
         "viridis","dplyr", "gdistance","vegan", "gridExtra",
         "psych", "iNext", "BAT", "ggplot2") # package names
# install.packages(pkgs)
# update.packages()
```

## ...and load packages

```{r, warning = FALSE}
library(here) #for easy directory management

# soon not going to be supported any more
library(sp) 
library(raster)
library(rgdal)
library(rgeos)
library(rgl)
library(rasterVis)

# the soon up-to-date spatial packages
library(sf)
library(tmap)
library(dismo)
library(GISTools)
library(maptools)

library(viridis) # nice colour palettes
library(units)
library(ggplot2)
library(gridExtra)
```

## How to call help 

```{r}
#- Information about a function or a package. If you do not understand some
# of the code below, always check the arguments and help of the function.
?dismo
#- search for an item:
??SpatialPolygons
#- show the instructions of a package:
vignette(package = "dismo") # vignette(package="sp")
#- load the instruction of a package (embedded pdf):
vignette("over") #vignette("sdm")
```


```{r, eval=FALSE, echo=FALSE}
sp::plot
showMethods(f = 'plot') #S4 type method
methods(generic.function = 'plot') #S3 type method
```

## Assign the workspace

Set your root directory, e.g. the directory under which you store everything
you need for this course, i.e. the data and the R-Scripts. You could name it
'TUB_BD_Course'.

Create the subfolders relative to root_wd. Please adjust to your setup. A possible
folder structure could be: 
<br>

-- TUB_BD_Course
<br>
----- R          # store here all your scripts, i.e. this script day_2.R,..
.<br>
----- _Data_Borneo    # the Borneo data 
<br>
-------- BaseMaps      # geo data, found in _Data_Borneo
<br>
-------- BaseRecords   # observation data, found in _Data_Borneo 
<br>

<br>

```{r}
# adjust the working directory [wd]
here()   # where are you? -> you should go to your root/ parent directory
getwd()  # alternative fct

# Set your OWN root directory, the following is just an example of how to do it:
root_wd <- setwd("C:/Users/kramer/PopDynIZW Dropbox/Steph Kramer/_GitHub")
```
<br>

Now, we assign the path to these folders relative to the root directory:
<br>

-- root_wd      <br>
----- R           <br>
----- dbor_wd    <br>
-------- maps_wd    <br>
-------- recs_wd    <br>


```{r}
#today, we start with Borneo
dbor_wd     <- paste0(root_wd, "/", "_DataBorneo") # mind difference paste0() and paste()
  maps_wd   <-  paste(dbor_wd, "/", "BaseMaps",    sep = '')
  recs_wd   <-  paste(dbor_wd, "/", "BaseRecords", sep = '')
```
<br>
Create an output folder where you can temporarily store your results (anything
you plot, create, want to save,....) with *dir.create*.

```{r}
output_wd <- paste0(dirname(root_wd), "/", "_A_Borrar")
if (!dir.exists(output_wd)) {dir.create(output_wd)}

```
Finally, your folder structure will look like this:

-- TUB_BD_Course <br>
----- R          # store here all your scripts, i.e. this script day_2.R,...
<br>
----- output     # for everything you create/ plot/ save
<br>
-------- _Data_Borneo    # the Borneo data
<br>
-------- BaseMaps      # geo data, found in _Data_Borneo 
<br>
-------- BaseRecords   # observation data, found in _Data_Borneo 
<br>

<br><br>


## Access folder content

You will find a lot of bioclimatic raster files (.asc) as well as shapefiles in the folder
```{r}
filenames <- list.files(path=maps_wd)
##run also :
# filenames
head(x = filenames)
```
<br>

## Recap

**useful functions:**
<br>
*head()* : only shows the first 6 elements, i.e. if you have a large data frame, you can quickly check header and the first entries (analoguously: tail() )
<br>
*paste(   , sep='')* : appends characters or numbers - great for working directories, loops across many maps or data.frames etc
<br>
**data formats and access:**
<br>
*data.frame*  :  [row, column] or "dollar sign", i.e. mydata[1,2]: first row, second column, or     mydata$MyColumnName[1]
<br>
*matrix*  :  see above
<br>
*list*    :  [[element]], i.e. mylist[[1]]. Achtung! A list element can consist of a data.frame, that you can then access like that: mylist[[1]][1,2]....
<br>
*S4-objects* (like spatial data)  :  @, i.e. RasterLayer@data@values


<br><br>

# Data Types 

##  Raster data

### Data import 

First, load a raster map. Please check the description in the folder called
DataDescription_Borneo.doc. The file bio_asc_01.asc is showing mean annual 
temperature * 10. Why?  Because it can be stored as integer, not floating type, 
which saves a lot of storage and memory.
```{r}
# read the ascii file as raster format
ras1 <- raster(x = paste0(maps_wd, '/bio_asc_01.asc'))
Bor_mat <- ras1 #easy copying of whole maps
```
<br>
Now have a look at the object
```{r}
ras1
```

What does the slot with extent tells you? Looks like decimal degrees, i.e. could be
latitude and longitude. However, the slot crs (crs stands for coordinate reference system)
is empty. That means, without knowing the reference system, the map cannot be
plotted at the exact location on the globe. Thus, when you get a map, always
make sure you also get the information of the coordinate reference system! If it 
is not yet set (because you import a raster map as a simple matrix or points from
a file), this is the first thing you should do!
Btw - you can access the information of an S4-class such as RasterLayer objects (https://rdrr.io/cran/raster/man/raster.html) with '@':

```{r}
# str(ras1) #check this
ras1@crs    # crs = coordinate reference system. If empty, assign it!
```
<br>

### Assign coord. ref. system 

Important! Please read and repeat basic GIS-knowledge on coordinate reference 
systems (crs), map projections, and geographic and projected coordinates.  

If a map already has a defined crs, then you cannot simply overwrite this! To
transform the coordinates, e.g. angular coordinates of latitude/ longitude of WGS84, 
into a projected crs with planar coordinates (e.g. equal area of Lambert Azimuthal),
you need to *transform* it (see below chapter **Raster Projection**).

<br>
In the following, we assign the crs only when no information is provided. This can be 
the case when you upload a map based on a simple data frame or matrix, i.e. when
it is not a spatial object.  
@crs gives the information of the object (here: raster layer) Bor_mat or ras1, and
CRS() is calling to a function that sets/defines the crs of the object. 
Mind the spelling in small letters/ capitals! 
crs = shows coordinate reference system of a spatial object;
CRS creates projection and sets arguments for crs!

An easy way to specify the crs is using the EPSG code, e.g. instead of
CRS("+proj=longlat +datum=WGS84") use 
CRS("+init=epsg:4326").

See http://spatialreference.org/. The code for WGS84 is 4326.


```{r}
ras1@crs <- CRS(projargs = "+proj=longlat +datum=WGS84")# equal to
ras1@crs <- CRS("+init=epsg:4326")
# mind: crs != CRS
```
<br>
Projections can also be transferred from one object to another, 
but only when the crs hasn't yet been specified, i.e. when the @crs -slot is empty.
In the following, load the topographic map (digital elevation model) of 
Borneo (Bor_dem.asc or bio_asc_24.asc) and assign it the same crs as Bor_mat, 
because we know they are overlaying (you can check that with the raster extents):
```{r}
# ras24 = DEM = digital elevation model
ras24 <- raster(x = paste0(maps_wd, '/bio_asc_24.asc')) 
ras24@crs <- ras1@crs # same as: CRS("+proj=longlat +datum=WGS84")
```
<br><br>

##### Plot for a quick look

There are different ways to plot a raster map. Below is a chapter on visualising
raster data, here I just give a very quick overview to get a first
impression of the data we work with.

The easy base plot:
```{r}
# base plot
plot(x = ras1)
```
<br>
Using the tmap package:

```{r}
# static plot
tmap_mode(mode = "plot")
tm_shape(shp = ras1) + tm_raster()
```
<br>
Very good for checking the correct location of the data. Click on the + | - | layer 
icons on the upper left and zoom in or out and change the background layers:

```{r}
# interactive plot
tmap_mode(mode = "view")
tm_shape(shp = ras1) + tm_raster()
```
<br><br>

### Clip areas

We do this first to work with a small raster to save computation time. For this, 
we create a clip_extent based on the spatial coordinates. The command 'crop' then 
clips the raster map.
```{r}
extent(x = ras1) # same as extent@ras1
clip_extent <- extent(117.2, 117.45, 5.4, 5.5)
ras1_cr <- crop(x = ras1, y = clip_extent)
plot(ras1_cr, col= inferno(10))
```
<br>

### Accessing *RasterLayer*s

First, have a look at the internal data structure
of the raster object:

```{r}
ras1_cr # str(ras1_cr) # attributes(ras1_cr) # class(ras1_cr)
```
<br>
There are many ways to retrieve internal data and access single bits of 
information:
```{r}

bbox(obj = ras1_cr)
ras1_cr@extent@xmin
   # xmin(ras1_cr) # xmin(extent(ras1))
   # ras1_cr@extent # extent(ras1_cr) # attr(ras1_cr,'extent')
ras1_cr@ncols        # ncol(ras1_cr) !n.b. cols vs. col!
head(ras1_cr@data@values)  # values(ras1_cr)
ras1_cr@crs
   # crs(ras1_cr) # projection(ras1_cr) !=CRS(ras1_cr)

# crs = coordinate reference system defined
# CRS creates projection and takes args for crs!
# e.g. CRS("+proj=longlat +datum=WGS84")
```


```{r}
#Retrieve internal data cont.:
nrow(x = ras1_cr) # ncol(ras1_cr) # ncell(ras1_cr)
dim(x = ras1_cr)  # 12 rows, 30 columns, 1 z-dimension
res(x = ras1_cr)  # resolution = cell size
```
<br>

An important function is *coordinates()*, which gives you the centroids of
each raster cell.
```{r}
head(x = coordinates(obj = ras1_cr), n = 10) # command from package 'sp'
```
<br><br>

### Terrain computation

We will now work with the digital elevation model (DEM) ras24 and create
a hillshade (3D look) based on topography using slope and aspect:
- simulates a 3D surface
- computes shaded relief values for a raster surface
```{r}
slope <- terrain(x = ras24, opt = "slope",
                    unit = "radians", neighbors = 8)
aspect <- terrain(x = ras24, opt = "aspect",
                     unit = "radians", neighbors = 8)
Bor_hs <- hillShade(slope, aspect, angle = 45,
                      direction = 270)
```
<br>
Plot the hillshade
```{r}
plot(Bor_hs, col=grey(0:100/100), legend=FALSE)
```
<br>
Other very useful terrain calculations are cost surfaces, cost distances
and least cost path, e.g. for corridor calculations. We will do that in
another course.
<br><br>

### Visualising rasters

Difference between *plot* and *image*

#### plot

*plot* keeps the proportion of the map/ cells (e.g. squares here).
```{r}
plot(x = ras1, col=viridis(256))
#plot(ras1_cr)
```
<br>
However, if you plot several layers on top of each other, *plot* will not 
overlay them exactly, because the argument 'add' fails (example below). 
In that case, better use *image*, or any other plot function.

#### image (raster)

If you use *image*, the plotted proportions will be changed and skewed to the
plot surface.
```{r}
image(x = ras1)
# image(ras1_cr)
```

<br><br><br>

#### tmap (interactive plots)

you can also save the interactive map - check your output folder for the html 
file. Tip of the day: always save the epsg code of your crs at the end of
a spatial file.
```{r}
tmap_mode("view")
(i.m <- tm_shape(ras1) +
  tm_raster(palette = "viridis",  title = "Mean annual temp"))
tmap_save(i.m, paste0(output_wd,"/BorneoMAT_4326.html"))

```
<br>
You can also use this package to create high quality plots for your publications.
```{r}
tmap_mode("plot")
(m <- tm_shape(ras1) +
  tm_raster(palette = "viridis",  title = "Mean annual temp"))
tmap_save(m, paste0(output_wd,"/BorneoMap_4326.png"), 
          units = "mm", width = 70, height = 70, dpi = 150 )

```
<br><br>

#### ggplot with raster

For advanced R-users : Beautiful plots can also be made with ggplot2; however, 
RasterLayer objects cannot directly be plotted, but have to be converted to a 
conventional data frame first, i.e. 2 steps are needed prior to plotting. The
object SpatialPointsDataFrame is explained in more detail when working with
vector data.

```{r}
# First, convert RasterLayer object to a SpatialPointsDataFrame
ras1_pts <- rasterToPoints(ras1, spatial = TRUE)
# Then to a dataframe
ras1_pts  <- data.frame(ras1_pts)

ggplot() +
  geom_raster(data = ras1_pts , aes(x = x, y = y, fill = bio_asc_01)) + 
  scale_fill_gradientn(colors=viridis(100),name = "°C * 10") +
  xlab("Longitude") +
  ylab("Latitude") +
  ggtitle("mean annual temperature")
```
<br><br>

#### composite plots

##### Simple composite plot

Plot the hillshade (3D relief) and add temperature colours on top: alpha value 
gives semi-transparency. 

```{r}
image(Bor_hs, col=grey(0:100/100))
image(ras24, col=terrain.colors(25,alpha=0.3),add=T)
points(coordinates(ras1_cr),cex=0.1,pch='+')
plot(extent(ras1_cr),add=T, col='red')
```
<br>
Now open the image-plot with the zoom icon in a separate window and change the size.
You will see that the different layers are still plotted on top of each other.
<br>
Now do the same with the base plot function:
```{r}
plot(Bor_hs, col=grey(0:100/100),legend = FALSE)
plot(ras24, col=terrain.colors(25,alpha=0.3),add=T)
points(coordinates(ras1_cr),cex=0.1,pch='+')
plot(extent(ras1_cr),add=T, col='red')
```

The extent is plotted as a red box, and the centroid coordinates of the raster
cells are plotted as points. Use the 'zoom' function to increase the plot. To
do so, call 'zoom' and click twice on the map to span your rectangle. Check out 
the area around the red box:
```{r}
# zoom()
# locator(2) # gives you the coordinates of - here 2 - locations interactively
```

<br> <br>

##### composite plot with tmap

In the following, we will create a SpatialPointDataFrame (sf-package) from the
small clipped/ cropped raster and plot it on top of the hillshade. Use the
possiblitiy of selecting/ unselecting different layers (layer icon on the left).

```{r}
hillsh <- Bor_hs

# make sf object from coordinates -> you will learn that also later
ras1_cr_sf <- st_as_sf(data.frame(coordinates(ras1_cr)), # create dataframe of coordinates
                       coords = c("x", "y"), # define columns for the coordinates
                       crs = 4326, # define crs, 4326 is the EPSG code
                       sf_column_name = "geometry") # sf needs a geometry column and you have to name it


# interactive plot
tmap_mode(mode = "view")
tm_shape(shp = hillsh, raster.downsample= TRUE)  +  
         tm_raster(palette = "Greys") +
  tm_shape(shp = ras24, raster.downsample= TRUE) + 
         tm_raster(palette = grDevices::topo.colors(20),alpha = 0.3) +
  tm_shape(shp = ras1, raster.downsample= TRUE) + 
         tm_raster(palette = grDevices::rainbow(10), alpha = 0.3) +
  tm_shape(shp = ras1_cr_sf) + tm_dots(shape = 20, size = 0.01) 
```

What do the shape-arguments mean, e.g. shape = 20? Have a look here at the symbols:
https://ggplot2.tidyverse.org/articles/ggplot2-specs.html?q=shape#sec:shape-spec
<br>
Unfortuantely, in tmap you can only use shape = 1 and shape =20, only circles are plotted
<br>


```{r}
# static plot  
tmap_mode(mode = "plot")
tm_shape(shp = hillsh, raster.downsample= TRUE) + 
  tm_raster(palette = "Greys") +
tm_shape(shp = ras24, raster.downsample= TRUE) +
  tm_raster(palette = grDevices::terrain.colors(10), alpha = 0.3) +
tm_shape(shp = ras1_cr_sf) + tm_dots(shape = 1, size = 0.05) 
```
<br> <br>


#### perspective: 3D plot

```{r}
#- Cool 3D plots with rgl library, e.g. 'rgl.surface'
persp(x = ras24, xlab = "Easting", ylab = "Northing",
      zlab = "elevation", r = 5, d = 1.5, expand = 0.1,
      ticktype = "detailed")
```
<br>

Why is it so spiny??? Check units! This is still a geographic crs, not a
projected one. Wait for the lecture for details.

<br><br>

### Raster projection

Project the Grid  (the small one!!!) to have all units in meters!
```{r}
proj_moll=CRS("+proj=moll +lat_0=65 +lon_0=10") # Mollweide
Bor_dem_moll <-projectRaster(ras1_cr,crs=proj_moll)
persp(Bor_dem_moll, xlab="Easting", ylab="Northing",
      zlab="elevation", main="Elevation model of Borneo",
      r=1, d=5.5, expand=0.1, ticktype="detailed")
```
<br><br>

### Raster Stacks

A raster stack is a collection of RasterLayer objects
with the same spatial extent and resolution, similar to a geodatabase. 
Go into the maps-folder
and check what is inside:

```{r}
# gives names and full path of file
files.full <- list.files(path = maps_wd, pattern = '.asc$',
                         full.names = TRUE )
# files.full # check also 
files.full[1:2]

# names only
files.rel <- list.files(path = maps_wd, pattern = '.asc$',
                        full.names = FALSE )
files.rel[1:2]
```

<br><br>

#### Working with stacks 

The advantage is, that you do not need to apply a command to each raster map 
separately, but can do it 'all in one'. E.g., We can set the crs of each 
single raster in just one line. In the following, we *stack* four maps in
a 'geodatabase' called 'predictors'. Check the .doc for a description of
the layers. In the following, we select 4 spatial layers:

```{r}
predictors <- stack(x = files.full[c(9, 12, 22, 24)])
projection(x = predictors) # crs(predictors)
projection(x = predictors) <- CRS(projargs = "+proj=longlat +datum=WGS84")
```
<br>
A raster stack contains the single raster layers in a list
```{r}
predictors@layers # this is a list! Address single layers with [[ ]]
```
<br>
Plotting just one layer:
```{r}
plot(predictors@layers[[4]])
```

<br><br>

#### Plotting stacks 

With base plot

```{r}
plot(x = predictors)
```
<br>

With tmap: 
This is standardizing the legend to the min and max of all layers, which is 
problematic if maps are not in the same units. That's why you cannot see the
information in two of the maps:
```{r}
tmap_mode(mode = "plot")
tm_shape(shp = predictors) + tm_raster()
```
<br>
<br>
<br><br><br><br><br>

#### Extract values from stacks

Violinplots for summary statistics
```{r}
bwplot(x = predictors[[c(1, 2)]]) # bwplot(predictors)
# n.b. double [[ ]] because stack is a list of spatial rasters
```
<br>
Again, these do not look nice becuase of the different axis scaling.
<br>
For advanced users: More beautiful violinplots can be found in the YaRrr package 
(called pirate plots: https://bookdown.org/ndphillips/YaRrr/pirateplot.html)
or simply in ggplot2. For this, we need to transform the RasterLayer data 
into a data.frame first:
```{r}
raster_df <- na.omit(data.frame(values(predictors[[c(1,3)]])))
raster_names <- names(raster_df)
raster_ct    <- dim(raster_df)[1]
df2 <- data.frame(val = c(raster_df[,names(raster_df)[1]], 
                          raster_df[,names(raster_df)[2]]))
df2$grp <- rep(raster_names , each = raster_ct)
head(df2)

#take a random subsample of the data to not crash your PC when plotting:
a <- sample(x=nrow(df2), size=1000, replace = FALSE) 
df3 <- df2[a,]
```

```{r}
p <- ggplot(df3,aes(x=grp,y=val))
p + geom_violin(scale = "width",fill = "grey80", colour = "#3366FF") + 
    geom_boxplot(width=0.2) + 
    geom_jitter(height = 0, width = 0.05, alpha=0.2, size=3,colour='blue')
```
<br>
Since the untis are so different, it makes more sense in this case 
to plot them separately, using include 'facet_grid' and scales = free.

```{r}
p <- ggplot(df3,aes(x=grp,y=val))
p + geom_violin(scale = "width",fill = "grey80", colour = "#3366FF") + 
    geom_boxplot(width=0.2) + 
    geom_jitter(height = 0, width = 0.05, alpha=0.2, size=3,colour='blue') +
    facet_grid(grp ~ ., scales = "free")  

# save the last plot
ggsave(paste0(output_wd,"/savedggplot.pdf"),width = 5, height = 5)

```

<br>
...or plot them separately using grid.arrange from package *gridExtra*:
```{r}
df3_bio9 <- subset(df3, df3$grp == 'bio_asc_09')
df3_bio24 <- subset(df3, df3$grp == 'bio_asc_24')

p1 <- ggplot(df3_bio9,aes(x=grp,y=val)) +
      geom_violin(scale = "width",fill = "grey80", colour = "#3366FF") + 
      geom_boxplot(width=0.2) + 
      geom_jitter(height = 0, width = 0.05, alpha=0.2, size=3,colour='blue') 
p2 <- ggplot(df3_bio24,aes(x=grp,y=val)) +
      geom_violin(scale = "width",fill = "grey80", colour = "#3366FF") + 
      geom_boxplot(width=0.2) + 
      geom_jitter(height = 0, width = 0.05, alpha=0.2, size=3,colour='blue')

grid.arrange(p1, p2, nrow = 1)
```
<br>
check here for how ggplot2 works if you want to make really nice plots:
*https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf
<br>
or get inspiration here:
<br>
*https://www.cedricscherer.com/2019/08/05/a-ggplot2-tutorial-for-beautiful-plotting-in-r/
<br>
*https://www.r-graph-gallery.com/
<br><br><br>

Back to spatial R.
<br>
Retrieve metrics/ statistics from the rasters:
```{r}
#- Extract information from all rasters in one command
# cellStats(predictors, 'mean')
round(x = cellStats(x = predictors, stat = 'mean'), digits = 2)
```
<br><br>

### Data export

Save the raster (not the plot...): There are many exchange formats for rasters. 
The best would be GeoTiff, which also saves the projection and is smaller than 
ascii. However, for modelling
e.g. in MaxEnt, the raster maps are needed in ascii-format.
```{r}
writeRaster(x = hillsh, 
            filename = paste0(output_wd,"/hillshade.asc"), 
            format = "ascii",
            overwrite = TRUE, 
            NAflag = -9999)
```
<br><br>

### Manipulating rasters

Do any kind of raster algebra:
```{r}
new_ras <- ras1 + ras24 + 100
```
<br>

For changing cell size use aggregate() or resample():
```{r}
# collapse 20*20 cells into 1 using function 'mean':
ras1_agg <- aggregate(x=ras1,fact=20, fun= mean) 
```

<br>

We finally plot the two new rasters next to each other:

```{r}
par(mfrow=c(1,2)) # recap from course 1: plots in 1 row and 2 columns
plot(x = new_ras, col = rev(rainbow(5)))
plot(ras1_agg)
par(mfrow=c(1,1)) # remember to set it back to the default
```

<br><br>

#### Change and query raster values

```{r}
# convert ras1 to degree Celsius units (divide by 10)
range(values(ras1), na.rm = TRUE)
mean.t.c <- ras1 / 10
range(values(mean.t.c), na.rm = TRUE)
# find areas with mean annual temp >= 25 deg C
mean.t.c.25 <- mean.t.c >= 25
```
<br><br>
 Have a look at the binary plot

```{r}
plot(x = mean.t.c.25)
```
<br><br>

#### Accessing cell values

Remember the function **extract()**. In the following, we extract from the
small raster the first three values (= col 1-3 ; longitude or x-axis) 
at line 5 (= row 5 ;latitude or y-axis).

```{r}
cells <- cellFromRowCol(object = ras1_cr, row = 5, col = 1:3)
cells    #gives cell ID number, the index/ rownumber!!!
extract(x = ras1_cr, y = cells) # gives cell values!
plot(x = ras1_cr)
# to plot the points on top, insert the 'cells' index into 
# the dataframe of the coordinates of the RasterLayer object ras1_cr
points(x = coordinates(obj = ras1_cr)[cells,], col = 'blue')
```
<br><br>
   
Extract rastervalues from all rasters at once (raster stack):
<br>
This is very important if you want to create your master table, i.e. based 
on the xy-coordinates of your species sightings (ie. GPS point locations), 
you could extract the important environmental predictors at that location. 
<br>
We do that for an area in the center of Borneo. For this, we first get the
row and column indices of the center of the large Borneo map, e.g. ras1, and
5 x 5 cells in addition, i.e. a square with an area of 5 km x 5 km :
```{r}
center_x = floor(nrow(predictors)/2) # learn about the functions round(), 
center_y = floor(ncol(predictors)/2) #                 ceiling(), floor()
center_x;center_y
stack_cells <- cellFromRowCol(object = ras1, row = center_x:(center_x+5), 
                                             col = center_y:(center_y+5) )
head(stack_cells) 
```
<br>

Why are these index numbers so large? Because the counting
starts at the upper left corner of the map and continues consecutively, i.e.
the data (ras1@data@values) are one huge vector with length nrow * ncol !
<br>

```{r}
# finally, extract them:
pred_dat <-   extract(predictors, stack_cells) 
head(pred_dat)
mastertable <- data.frame(stack_cells,pred_dat)
mastertable
```

<br> <br><br><br>

Change raster values:

```{r}
xy = xyFromCell(object = ras1_cr, cell = cells) # coordinates(ras1_cr)[cells,]
extract(x = ras1_cr, y = xy)          # extract(ras1_cr, cells)

# Change values, e.g. for adding forest or creating a corridor
# take care! irreversible! better work on a copy!
copy_ras <- ras1_cr
copy_ras[cells] <- 250
plot(x = copy_ras)
```
<br><br>

#### Compute distance to points

```{r}
# calculate distance
my_dist <- distanceFromPoints(object = ras1_cr,xy = xy)
plot(x = my_dist) #units?

```

```{r}
# nicer plot adding the points from which a distance should be computed.
xy_sf <- st_as_sf(x = data.frame(xy), 
                  coords = c("x","y"),
                  crs = 4326,
                  sf_column_name = "geometry" ) 

tmap_mode(mode = "plot")
tm_shape(shp = my_dist) + tm_raster(n = 100,
                                    palette = rev(grDevices::terrain.colors(100)),
                                    legend.show = FALSE) +
  tm_shape(shp = xy_sf) + tm_dots(size = 1)
```
<br><br>

   
Calculate distance from many points
```{r}
cells1 <- c(cells, 250, 360) # add two more points
xy=xyFromCell(ras1_cr, cells1)# =coordinates(Bor_mat_cr)[cells,]
my_dist <- distanceFromPoints(ras1_cr,xy)
plot(my_dist) #units?
```
<br><br>


## Vector data / shapefiles

### Polygons and lines

#### Import shapefiles

Currently, there are two packages *sp* and *sf* (standing for simple features),
both with still important functionality. However, sf is much easier to use and handle. 
The suggestion currently is: learn both!
https://www.r-bloggers.com/2018/03/should-i-learn-sf-or-sp-for-spatial-r-programming/
In the following, the most important commands are provided for the sf-package,
and if necessary, also for the sp-package:
```{r}
# Border of countries and provinces of Borneo
# - only loading columns 1:3, 5, 7, 17, 18 of attribute table:
Borneo_shp <- st_read(dsn = maps_wd,
                      layer = "borneo_admin",
                      stringsAsFactors = FALSE)[, c(1:3, 5, 7, 17, 18)]
# Protected areas (National Parks, Nature Reserves, Forest Reserves)
PA_shp     <-  st_read(dsn = maps_wd,
                   layer = "Bor_PA",
                   stringsAsFactors = FALSE)[, c(1:4)]
# Main rivers
River_shp  <- st_read(dsn = maps_wd,
                     layer = "sn_100000",
                     stringsAsFactors = FALSE)

## the 'old' sp package would work with that command:
Admin_shp <- readOGR(dsn=maps_wd, layer="borneo_admin",
              stringsAsFactors=FALSE)[,c(1:3,5,7,17,18)]
## transformations
Admin_sf <- as(Admin_shp, "sf") ## from sp object to sf object 
Admin_sp <- as(Admin_sf, "Spatial") ## from sf to sp object
```
<br><br>

#### Working with vector data

Accessing simple features - easy handling, similar to data.frames
```{r}
#Please note the similarity to accessing info from rasters.
str(object = extent(PA_shp))
extent(PA_shp)@xmin
names(x = PA_shp) #returns column names of a.t.
```
<br><br>

Accessing SpatialPolygonsDataFrames:
```{r}
# Much more complex than sf-objects.
class(Admin_shp)
str(object = extent(Admin_shp))
extent(Admin_shp)@xmin
names(x = Admin_shp) #returns column names of a.t.
```

<br><br>

Summarising spatial information for each column of attribute table [a.t.]:

```{r}
summary(object = PA_shp)
str(object = PA_shp)
PA_shp
attributes(x = PA_shp)
```
<br>

Have a look at the content

```{r}
head(x = PA_shp)   
tail(x = PA_shp)
```
<br>

Retrieving information of a.t. - recap working with data.frames 
from Day1_R-Intro course.

```{r}
PA_shp[1, ] #returns first entry (row) of all 4 columns
```

```{r}
PA_shp[, 2] # returns summary of col only!
```

```{r}
# PA_shp$NAME_ENG #returns a long list  
head(x = PA_shp$NAME_ENG) #using fct head() to only show the first 6 entries
```

```{r}
PA_shp[1, 1] # = PA_shp$SITE_ID[1]
PA_shp[2, 3] # = PA_shp$COUNTRY[2]
```

<br>

The following returns the row indices of the data frame or a.t., respectively:
```{r}
which(x = PA_shp$COUNTRY == 'Malaysia')
```

<br>

#### Visualising vector data

You can use many packages, e.g. ggplot2, for sf-objects:
https://www.r-spatial.org/r/2018/10/25/ggplot2-sf.html.
However, it does NOT work for sp-objects (SpatialPolygonsDataFrame ).
```{r}
ggplot(data = PA_shp) +
    geom_sf() + 
    xlab("Longitude") + ylab("Latitude") +
    ggtitle("Protected areas")
```
<br><br>

or tmap:
```{r}
tmap_mode(mode = "plot")
tm_shape(shp = PA_shp[, 1]) + tm_polygons(col = "SITE_ID",
                              palette = grDevices::terrain.colors(5),
                              style = "quantile") # color scheme by quantiles
```
<br><br>

Interactive mode
```{r}
tmap_mode(mode = "view")
tm_shape(shp = PA_shp[, 1]) + 
  tm_polygons(col = "SITE_ID",
                    palette = grDevices::terrain.colors(5),
                    style = "quantile") # color scheme  by quantiles
```
<br><br>

It is also possible to use a simple plot() function for sf-objects. However, mind that this
object has 7 columns, so each column holds information that will be plotted, i.e.
you will plot 7 single plots. So do a little transformation:
either plot just one column, or define it as spatial:

```{r}
# plot(Borneo_shp) not run
plot(Borneo_shp[,1]) # select column
```

```{r}
plot(as(Borneo_shp, "Spatial")) #plot geometry without information of a col
```


And for the sp-package:

Here, the basic plot function works:
```{r}
plot(Admin_sp)
```

There is also a special function for plotting sp-objects, however, also here, you
need to select which information you want to plot.
```{r}
# spplot(Admin_sp) #Don't do that! Each column of the a.t. will be plotted
head(Admin_sp)
spplot(Admin_sp[6]) # only plot one of the geometries, coloured e.g. by shape_length
```

Plot a detail on the map - Plot of a single selected protected area :
```{r}
tmap_mode(mode = "plot")
tm_shape(shp = Borneo_shp) + tm_polygons(border.col = "deepskyblue4") +
  tm_shape(PA_shp[,1]) + tm_polygons(border.col = "black") +
  tm_shape(PA_shp[1, ]) + tm_polygons(border.col = "red")
```

<br><br>

#### Small exercises

Simplify your life! Work with subsets, select Malaysia, save and plot as shapefile

```{r}
Mal_PA_shp <- subset(PA_shp, PA_shp$COUNTRY == 'Malaysia')
st_write(obj = Mal_PA_shp, 
         dsn = output_wd, 
         layer = 'test',
         driver = 'ESRI Shapefile', 
         delete_layer = TRUE)

tmap_mode(mode = "plot")
tm_shape(shp = Borneo_shp) + tm_polygons(border.col = "blue") +
  tm_shape(shp = Mal_PA_shp) + tm_polygons(col = "green")
```
<br><br>

Extract elevation per PA and save average to a.t. as a new column

```{r}
fewPA <- Mal_PA_shp[c(1:5), 1]
tmp <- extract(x = ras24, y = fewPA)   # returns a list - each element contains
                                       # the elevation raster cells (ras24)
str(tmp)
#lapply(tmp, FUN = summary)
```

```{r}
# Append mean elevation:
fewPA$mean_elevation <- round(x = unlist(lapply(tmp, FUN = mean, na.rm = T)))
fewPA
```
<br><br>

#### Geospatial calculation

Using the 'old' sp-package and package rgeos, commands for calculating 
properties of spatial objects were: gArea, gLength, gDistance. 
Other often used GIS functions relate to basic spatial operations, such as
gBuffer, gCentroid, gDifference, gUnion, and gIntersection, 
gUnionCascaded (= dissolve function) and gSimplify.
Spatial queries were gIntersects, gContains, gIsValid.
http://www.nickeubank.com/wp-content/uploads/2015/10/
    RGIS2_MergingSpatialData_part2_GeometricManipulations.html. <br>

The new sf-package has the same functionality, with functions starting with 'st_',
e.g. st_buffer, st_intersection etc.<br>
https://cran.r-project.org/web/packages/sf/vignettes/sf1.html
<br>
https://geocompr.robinlovelace.net/spatial-operations.html
<br>


##### Spatial query

Returns the area in square meters
```{r}
# st_area(x = Borneo_shp) #returns long vector
head(st_area(x = Borneo_shp))

```
<br>

##### Spatial transformation (projection)

```{r}
Borneo_shp_moll <-  st_transform(Borneo_shp, c("+proj=moll +datum=WGS84"))
class(Borneo_shp_moll) #sf object, data.frame!
```
<br>

```{r}
tmap_mode(mode = "plot")
tm_shape(shp = Borneo_shp_moll) + tm_polygons(border.col = "blue") #do you see the difference?
head(st_area(x = Borneo_shp_moll)) # units?
```


<br>

Example: Area of Malaysia in Borneo.

```{r}
Mal_Borneo_shp <- subset(Borneo_shp_moll, Borneo_shp_moll$NAME_0 == 'Malaysia')
head(st_area(x = Mal_Borneo_shp)/1000000) # or / 1e6
# or use set_units to change the units from m^2 to km^2
head(set_units(x = st_area(x = Mal_Borneo_shp), value = km^2) ) 
```
<br>

Example: Calculate area of a polygon and add the area of a polygon as new column to a.t.
```{r}
st_area(x = Borneo_shp_moll[3, ]) # for a single polygon
head(st_area(x = Borneo_shp_moll, byid = TRUE)) #for all polygons

area_km2 <- set_units(x = st_area(x = Borneo_shp_moll, byid = TRUE), value = km^2)
Borneo_shp_moll = data.frame(Borneo_shp_moll, area_km2)
head(x = Borneo_shp_moll)
```
<br><br>
The following for sp objects:  
```{r}
#gArea(Borneo_shp_moll, byid=TRUE)/1e6 # does not work, as Borneo_shp_moll is sf-object!
head(gArea(Admin_sp, byid=TRUE)/1e6) # what does the error message mean? -> check crs!
Admin_sp_moll <- spTransform(Admin_sp, c("+proj=moll +datum=WGS84"))
head(gArea(Admin_sp_moll, byid=TRUE)/1e6) # now it works!
```
<br><br>

### Points

#### Import spatial points

Usually, shapefiles come with a .proj file defining the projection, but not
always. Please always check whether the projection is set.
```{r}
pt_shp <- st_read(dsn = paste(recs_wd, 'FCsim.shp', sep = '/'))
pt_shp  # crs is missing!
st_crs(pt_shp) = 4326 # set it with command st_crs
pt_shp
```

Spatial points are defined by two columns with x and y coordinates, can be 
imported as normal data.frames and then converted to spatial objects.
```{r}
pt_file <- paste(recs_wd, 'MyNewSpecies.csv', sep = '/')
df_recs <- read.table(file = pt_file, header = TRUE, sep = ',')
class(x = df_recs)
head(x = df_recs)
```
See below of how to convert a data.frame into a spatial sf-object.
<br>

#### Plot the points

Simple data frames with location information, as long as the coordinates are the same, 
can also be plotted in space. We first plot the polygon
```{r}
plot(x = as(Borneo_shp, "Spatial"), col = 'grey', border = 'white') # polygon
points(x = df_recs$long, df_recs$lat, cex = 0.5, pch = 15) #simple d.f.! 
plot(x = as(pt_shp, "Spatial"), col = 'blue', add = T) 
```
<br>

Tmap can only handle spatial objects, so the data.frame df_recs needs to be
converted into an sf-object first. You have to define which columns contain the coordinates.
```{r}
recs_sf <- st_as_sf(x = data.frame(df_recs),
                       coords = c("long", "lat"),
                       crs = 4326,
                       sf_column_name = "geometry" )
```
<br>

Plot it with tmap.
```{r}

tmap_mode(mode = "plot")
tm_shape(shp = Borneo_shp) + tm_polygons(col = "grey",
                                         border.col = "white") +
  tm_shape(shp = recs_sf)  + tm_dots(shape = 3, size = 0.5) +
  tm_shape(shp = pt_shp)   + tm_dots(col = "blue") 
```
<br>

#### Spatial overlay - vector data

```{r}
# retrieve the geometry (location) indices of PA_shp at
# the locations of sp_recs: which points are in PA_shp
insidePA <- st_intersection(x = recs_sf, y = PA_shp) 
```
<br>

#### Spatial overlay - vector and raster

```{r}
# for a RASTER: extract mean ann. temp. from ras1
# and add it to a.t.
mean_t <- extract(x = ras1, y = recs_sf)
recs_sf$mean_t <- mean_t
mean(x = recs_sf$mean_t) # hist(sp_recs_sf$mean_t)

```
<br>

#### Saving vector data

```{r}
st_write(obj = insidePA,
         dsn = output_wd, 
         layer = 'inPA',
         driver ='ESRI Shapefile',
         delete_layer = TRUE)
```

<br><br><br>

At the end, a beautiful plot
```{r}
tmap_mode(mode = "plot")
tm_shape(shp = hillsh) + 
  tm_raster(palette = "Greys",
            legend.show = FALSE) +
  tm_shape(shp = ras24) + 
  tm_raster(palette = terrain.colors(25),
            alpha = 0.3,
            legend.show = FALSE) +
  tm_shape(shp = River_shp) + 
  tm_lines(col = "dodgerblue3") +
  tm_shape(shp = recs_sf) + 
  tm_dots(shape = 16, 
          size = 0.1) +
  tm_shape(shp = insidePA) + 
  tm_dots(col = "red",
          size = 0.5,
          shape = 3) +
  tm_shape(shp = PA_shp) + 
  tm_polygons(border.col = "white", 
              alpha = 0) +
  tm_layout(title = "END OF SESSION", 
            title.color = "darkgrey",
            title.position = c(0.01,0.9),
            title.size = 2)
```
<br><br><br>

# END
Very helpful when going back to old scripts or using scripts of others.
```{r}
sessionInfo()
```

